A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2024-01-29 pilot-publication/manuscript.Rmd
Work on manuscript code

<!-- Source main code to make global objects available in the text -->
```{r source, echo = FALSE, include = FALSE}
source("main.R")
```

@@

**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(n.m30d.atls))` (`r p.m30d.atls`%) patients died within 30 days in the ATLS arm, `r n.m30d.ptc` (`r p.m30d.ptc`%) patients in the PTC arm, and `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

- Cluster randomized controlled trial comparing the effect of ATLS, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated project officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.
-->
\newpage

@@

There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021], but there are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020].  We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS and PTC with standard care.

@@

The ATLS^®^ training was conducted in an ATLS certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018]. The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018]. These courses were conducted over a period of 2.5 to 3 days. The residents certified "pass" were considered as trained in respective courses.

@@

### Sequence generation
We used simple randomisation to allocate sites to trial arms. We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting<!-- Meeting notes ../meetings/20220307/ -->. The content of the envelopes dictated what trial arm (ATLS, PTC, or standard care) each hospital was allocated to. There were two hospitals in each of the intervention arms and three hospitals in the control arm.

@@

We compared participant outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care. In each combination we compared both differences in final values and differences in change from baseline. For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training. For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.

@@

Between `r start.date` and `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres. The ATLS arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients. We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS, and  `r n.ptc.residents` in PTC.

@@

```{r, echo = FALSE}
## Crude 30 day mortality
m30d.atls <- inline_text(post.training.characteristics.table,
  variable = outcomes__alive_after_30_days,
  level = "Yes",
  column = "ATLS"
)
m30d.ptc <- inline_text(post.training.characteristics.table,
  variable = outcomes__alive_after_30_days,
  level = "Yes",
  column = "PTC"
)
m30d.standard.care <- inline_text(post.training.characteristics.table,
  variable = outcomes__alive_after_30_days,
  level = "Yes",
  column = "Standard care"
)

## Extract tables for absolute and relative differences
standard.care.vs.atls.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]

standard.care.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]

atls.vs.ptc.table <- outcome.results.tables$ci.level.0.95[["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]

## Absolute differences
arr.standard.care.atls <- inline_outcome_text(
  standard.care.vs.atls.table,
  outcome = "30 day mortality",
  column = "Absolute difference",
  level = "Yes"
)

arr.standard.care.ptc <- inline_outcome_text(
  standard.care.vs.ptc.table,
  outcome = "30 day mortality",
  column = "Absolute difference",
  level = "Yes"
)

arr.atls.ptc <- inline_outcome_text(
  atls.vs.ptc.table,
  outcome = "30 day mortality",
  column = "Absolute difference",
  level = "Yes"
)

## Relative differences
rr.standard.care.atls <- inline_outcome_text(
  standard.care.vs.atls.table,
  outcome = "30 day mortality",
  column = "Relative difference",
  level = "Yes"
)

rr.standard.care.ptc <- inline_outcome_text(
  standard.care.vs.ptc.table,
  outcome = "30 day mortality",
  column = "Relative difference",
  level = "Yes"
)

rr.atls.ptc <- inline_outcome_text(
  atls.vs.ptc.table,
  outcome = "30 day mortality",
  column = "Relative difference",
  level = "Yes"
)

## Extract tables for change from baseline in 30 day mortality
absolute.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]
relative.change.from.baseline.table <- outcome.results.tables$ci.level.0.95[["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]

## Absolute change from baseline
abs.change.standard.care <- inline_outcome_text(
  absolute.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "Standard care",
  level = "Yes"
)

abs.change.atls <- inline_outcome_text(
  absolute.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "ATLS",
  level = "Yes"
)

abs.change.ptc <- inline_outcome_text(
  absolute.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "PTC",
  level = "Yes"
)

## Relative change from baseline
rel.change.standard.care <- inline_outcome_text(
  relative.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "Standard care",
  level = "Yes"
)

rel.change.atls <- inline_outcome_text(
  relative.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "ATLS",
  level = "Yes"
)

rel.change.ptc <- inline_outcome_text(
  relative.change.from.baseline.table,
  outcome = "30 day mortality",
  column = "PTC",
  level = "Yes"
)
```

After training, a total of `r n.m30d.control` (`r p.m30d.control`%) patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS arm and `r m30d.ptc` patients in the PTC arm. The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units. The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.

The absolute risk difference between the standard care and ATLS arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS and PTC arms (95% CI) `r arr.atls.ptc`% units. The relative risk in the standard care arm compared with the ATLS arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`. The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.

@@

# Discussion
We show that conducting a cluster randomized controlled trial
comparing ATLS with PTC and standard care is feasible in this setting. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables had very high missing data rates and may not be feasible to include in a full-scale trial.

@@

We found that the ATLS arm had lower 30-day mortality compared to the
PTC and standard care arms. We also found that the PTC arm had lower mortality than the standard care arm. These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS arm having the smallest sample size. This difference resulted from the randomisation process with a small number of heterogeneous clusters.

@@

Previous studies on the effect of in-hospital trauma life support
training on patient outcomes are observational or quasi-experimental
without a control group, with heterogeneous results
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008;
@Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
Cambodia found no significant effect on patient mortality after
implementing in-hospital trauma life support training
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008],
whereas one study from China that included 820 patients found a
significant reduction in mortality, from 20 to 15%, after implementing
ATLS [@Wang2010].

Several controlled trials, including at least two randomized
controlled trials [@Ali1996; @Ali1999], show that ATLS is associated
with improved provider skills [@Mohammad2013]. Observational evidence
indicates that PTC also leads to improved provider skills
[@Kadhum2020]. The missing link is then how, and if, these improved
skills translate into improved patient outcomes. As trauma care
providers we assume, and probably rightly so, that we deliver better
care if we train. The question is then how we should train, especially
considering the costs associated with some of the programs
offered.

>>>>

<!-- Source main code to make global objects available in the text -->
```{r source, echo = FALSE, include = FALSE}
source("main.R")
attach(results)
```

@@

**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(m30d.standard.care))` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

- Cluster randomized controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated project officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.
-->
\newpage

@@

There are observational studies indicating that trauma life support training programmes may improve patient outcomes [@Jin2021], but there are no randomised trials  [@Mohammad2013; @Jayaraman2014; @Kadhum2020].  We performed a pilot study aiming to assess the feasibility of conducting a cluster randomised controlled trial comparing ATLS^®^ and PTC with standard care.

@@

The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai, according to the standard ATLS^®^ curriculum [@acsAtls2018]. The PTC training was conducted in New Delhi, according to the standard PTC curriculum [@ptcfPtc2018]. These courses were conducted over a period of 2.5 to 3 days. The residents certified "pass" were considered as trained in respective courses.

@@

### Sequence generation
We used simple randomisation to allocate sites to trial arms. We prepared seven sealed envelopes from which one representative from each pilot site selected one in a common meeting<!-- Meeting notes ../meetings/20220307/ -->. The content of the envelopes dictated what trial arm (ATLS^®^, PTC, or standard care) each hospital was allocated to. There were two hospitals in each of the intervention arms and three hospitals in the control arm.

@@

We compared participant outcomes in three combinations of trial arms: ATLS^®^ versus PTC, ATLS^®^ versus standard care, and PTC versus standard care. In each combination we compared both differences in final values and differences in change from baseline. For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training. For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.

@@

Between `r start.date` and `r end.date`, we enrolled `r n.patients` trauma patients from `r n.centres` participating centres. The ATLS^®^ arm enrolled `r n.atls` patients, the PTC arm `r n.ptc` patients, and the standard care arm `r n.control` patients. We trained a total of  `r n.residents` residents, `r n.atls.residents` in ATLS^®^, and  `r n.ptc.residents` in PTC.

@@

After training, a total of `r m30d.standard.care` patients in the standard care arm had the primary outcome, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm. The absolute change from baseline in the primary outcome (95% CI) in the standard care arm was `r abs.change.standard.care`% units, in the ATLS^®^ arm `r abs.change.atls`% units, and in the PTC arm `r abs.change.ptc`% units. The relative change from baseline in the primary outcome (95% CI) in the standard care arm was `r rel.change.standard.care`, in the ATLS^®^ arm `r rel.change.atls`, and in the PTC arm `r rel.change.ptc`.

The absolute risk difference between the standard care and ATLS^®^ arms (95% CI) was `r arr.standard.care.atls`% units, between the standard care and PTC arms (95% CI) `r arr.standard.care.ptc`% units, and between the ATLS^®^ and PTC arms (95% CI) `r arr.atls.ptc`% units. The relative risk in the standard care arm compared with the ATLS^®^ arm (95% CI) was `r rr.standard.care.atls`, in the standard care arm compared with the PTC arm (95% CI) `r rr.standard.care.ptc`, and in the ATLS^®^ arm compared with the PTC arm (95% CI) the relative risk was `r rr.atls.ptc`. The absolute and relative differences in changes from baseline comparing the three arms are shown in Supplementary materials, along with data on all secondary outcomes.

@@

# Discussion
We show that conducting a cluster randomized controlled trial
comparing ATLS^®^ with PTC and standard care is feasible in this setting. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables had very high missing data rates and may not be feasible to include in a full-scale trial.

@@

We found that the ATLS^®^ arm had lower 30-day mortality compared to the
PTC and standard care arms. We also found that the PTC arm had lower mortality than the standard care arm. These findings indicate a large potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS^®^ arm having the smallest sample size. This difference resulted from the randomisation process with a small number of heterogeneous clusters.

@@

Previous studies on the effect of in-hospital trauma life support
training on patient outcomes are observational or quasi-experimental
without a control group, with heterogeneous results
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008;
@Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
Cambodia found no significant effect on patient mortality after
implementing in-hospital trauma life support training
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008],
whereas one study from China that included 820 patients found a
significant reduction in mortality, from 20 to 15%, after implementing
ATLS^®^ [@Wang2010].

Several controlled trials, including at least two randomized
controlled trials [@Ali1996; @Ali1999], show that ATLS^®^ is associated
with improved provider skills [@Mohammad2013]. Observational evidence
indicates that PTC also leads to improved provider skills
[@Kadhum2020]. The missing link is then how, and if, these improved
skills translate into improved patient outcomes. As trauma care
providers we assume, and probably rightly so, that we deliver better
care if we train. The question is then how we should train, especially
considering the costs associated with some of the programs
offered.
